**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach

### 2020-12-02 manuscript/ecosphere_resubmit_2/ecosphere_response_2.Rmd
Getting closer on edits.

@@

@@

@@

*Response:* Added in a new paragraph before the last paragaph of the discussion mentioning a few caveats.

>>>>

*Response:* You are correct.  This does assume that a change of the magnitude defined by the Nürnberg oligotrophic-mesotrophic cut-off would result in a change even for a lake starting with 0 chla, tn, or tp.  To clarify, we made a slight change to the wording on this and it now reads "... an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century."   The main change being "any oligotrophic lake" and added the following at the end: "A slope at or above these values would result in a oligotrophic-mesotrophic transition for all oligotrophic lakes, even lakes with the impossible situation of chlorophyll, total nitrogen, or total phosphorus concentrations of zero."

@@

*Response:* We agree that this paragraph was not focused and was also quite long.  We have split the paragraph apart.  It currently now has two paragraphs, with the first paragraph focusing on lengthening of the growing season and the second paragraph focusing on trophic state.  We also include your suggestions in this second paragraph.  The last part of our original paragraph that included other drivers and individual lakes was an admittedly ham-handed effort at covering our bases as well as dealing with reviewer comments.  Given that this last section of the paragraph does not add much to the discussion we have removed it.

@@

*Response:* Re-wrote this paragraph with a focus on these three main points.

@@

*Response:* Added in a new paragraph before the last paragraph of the discussion mentioning a few caveats.

### 2020-12-02 manuscript/ecosphere_resubmit_2/manuscript.Rmd
Getting closer on edits.

Add Paragraph about RI lakes: Context about human pressures on RI lakes + one sentence about adequate data being available

@@

For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.

@@

In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.  For example, changes in water quality at finer temporal scales, such as monthly or seasonal, may be different than the annual trends we analyzed.  Evidence in Rhode Island points to possible lengthening of the growing season as both May and October show increasing temperature trends over time [@noaatrends]. Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.  Additionally, the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.  For instance, are oligotrophic lakes showing different trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?  A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.  Lastly, there are other drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.  The analysis we report in this paper focuses on long term trends at broader scales.  It would be possible to use a similar approach to look at trends of individual lakes which could be a first step in trying to identify drivers of increasing chlorophyll.  These questions are beyond the scope of this study, but all warrant further investigation.

@@

Broad scale warming is difficult to manage on a sub-regional scale thus other mechanisms and approaches, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.  Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].  There are of course many other hypotheses that we could form through information obtained by long-term trends such as those reported here. Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs and turn that information into management action.

>>>>

<!-- Jeff start here for change to comment #4 -->
Add Paragraph about RI lakes: Context about human pressures on RI lakes + one sentence about adequate data being available

@@

For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century.  We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].  For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.  A slope at or above these values would result in a oligotrophic-mesotrophic transition for all oligotrophic lakes, even lakes with the impossible situation of chlorophyll, total nitrogen, or total phosphorus concentrations of zero.

@@

In addition to the sub-regional annualized trends of the five variables we address with this study, other trends may also be of interest.  For example, a lengthening of the growing season could increase the number of days with conditions that favor the growth of harmful algal blooms.  Many national and regional studies have documented longer growing seasons [e.g., @vega2019changes; @cooter1995recent; @kunkel2004temporal].  Evidence in Rhode Island also points to possible lengthening of the growing season as both May and October, the beginning and end of the sampling dates in our dataset, show increasing temperature trends over time [@noaatrends].  Furthermore, preliminary analysis of the URIWW data supports the idea that growing seasons may be getting longer in Rhode Island.  Cyanobacteria grow better at temperatures greater than than 25°C [@paerl2008blooms; @johnk2008summer; @reynolds2006ecology].  Early in our study period (1993-1995), the average first day of lake temperatures exceeding 25°C was June 11th, at the end of our study period (2014-2016) these warmer temperature were seen, on average, on June 1st.  A site-specific anomaly analysis of growing season length could shed further light on potential changes to growing seasons in Rhode Island lakes.

Trends, and their ecological importance, could also vary depending on trophic state of a given waterbody.  We identified a minimum ecologically important change of 0.035 µg/l per year of chlorophyll.  This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.  Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only a single rate for all lakes.  Furthermore, Stoddard et al. [-@stoddard2016continental] see an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.  A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al. [-@stoddard2016continental], that might be occuring only on oligotrophic lakes.   A full analysis and interpretation of trends by trophic state is beyond the scope of this study.  However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes. This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.

@@

Our analysis has shown warming lakes in Rhode Island, but broad scale warming of lakes has also been documented and shown to be a result of both climatic and local drivers and can vary greatly within regions [e.g., @oreilly2015warming].  Given that the drivers of warming are at both broad and local scales, suggests that managing warming lake temperatures will be a difficult task.  To counteract the impacts of continued warming on increased chlorophyll in lakes will, at a minimum, require additional interventions. Reducing nutrient loads below current levels is one such intervention and could be acheieved via source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].  In any event, controlling future eutrophication and protecting against harmful aglal blooms will continue to be a challenge against the backdrop of warming lakes.
